Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-07-28 Manuscript/proto-manuscript.tex
Methods partly updated

%The lack of electronic polarizability from the classical MD simulation
%models is a potential source of artefact, which could lead to overbinding
%of cations. 
%Almost all lipid bilayer simulations reported in the literature 
%are done with force fields that represent interactions between atoms
%with pairwise additive empirical potential and exclude the electronic 
%polarizability \cite{SOME_review, FF_papers}.
%The issue has been considered higly relevant since the early days of
%lipid bilayer simulations and pioneering simulation studies scaled
%the partial charges to effectively include electronic
%polarizability \cite{jonsson86,egberts94}.

@@

\subsection{Electronic continuum correction for lipid bilayers}
The explicit inclusion of
electronic polarizability has been, however, turned to be 
practically complicated and implementations for lipids are rare \cite{chowdhary13}.

According to the electronic continuum correction (ECC)\cite{leontyev11}, electronic
polarizability can be included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}. % that exists in living cells. 
Such a continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of ion force fields \cite{kohagen14,kohagen16,??}.

Here we apply the electronic continuum correction to lipid bilayers to accurately 
describe the lipid headgroup response to Na+ and Ca2+ concentrations \cite{catte16}. 
We use the Lipid14~\cite{dickson14} force field parameters as a starting point,
because they give the most realistic headgroup response with added cations 
and relatively realistic glycerol backbone and headgroup structures
when compared with other state of the art lipid models \cite{botan15,catte16}.
The partial atomic charges in Lipid14 were derived by fitting the electrostatic 
potential to its model quantum chemistry representation (RESP\cite{RESP_paper}) in vacuum.
If the charges are obtained by using RESP in an implicit solvent, 
they vary the most for the polar moieties \cite{maciejewski14}. 
In order to implicitly capture solvent induced polarization 
in models with fixed charges, 
we shall use the average of partial charges 
from both environments (i.e. vaccum and solvent),
so called implicitly polarized charges (IPolQ) \cite{ipolq2013}. 
By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
we represent the IPolQ charges in the electronic continuum correction of the Lipid14 model
by increasing the scaling factor $f_q$ to $0.8$. 
\footnote{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway, I use 0.80 as an approximate round value. The use of 0.78---.79 might be more appropriate, though.}

>>>>

@@

\subsection{Electronic continuum correction for lipid bilayers}
The lack of electronic polarizability from the classical MD simulation
models has been considered higly relevant issue since the early days of
lipid bilayer simulations and pioneering simulation studies scaled
the partial charges by half to effectively include electronic
polarizability \cite{jonsson86,egberts94}. Also approaches that
explicitly include electronic polarizability has been introduced, but
this has turned out to be practically complicated \cite{chowdhary13}.
In this work we effectively include electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. The approach is more sophisticated version of
early studies with scaling to half was used \cite{jonsson86,egberts94},
but significantly more simple than the explicit inclusion of electronic
polarizability \cite{chowdhary13}.

%is a potential source of artefact, which could lead to overbinding
%of cations. 
%Almost all lipid bilayer simulations reported in the literature 
%are done with force fields that represent interactions between atoms
%with pairwise additive empirical potential and exclude the electronic 
%polarizability \cite{SOME_review, FF_papers}.

According to ECC, electronic
polarizability can be included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}.
Such a continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of ion force fields \cite{kohagen14,kohagen16,??}.

The main goal of this work is to apply ECC correction to accurately 
describe the lipid headgroup response to Na$^+$ and Ca$^{2+}$ concentrations
in MD simulation models when compared with NMR data \cite{catte16}. 
The Lipid14~\cite{dickson14} force field parameters were used as a starting point,
because their response to bound ions was most realistic against NMR data
(see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and headgroup structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.

The partial atomic charges in original Lipid14 parametrization were derived by fitting the electrostatic 
potential to its model quantum chemistry representation (RESP\cite{RESP_paper}) in vacuum.
If the charges are obtained by using RESP in an implicit solvent, 
they vary the most for the polar moieties \cite{maciejewski14}. 
In order to implicitly capture solvent induced polarization 
in models with fixed charges, 
we shall use the average of partial charges 
from both environments (i.e. vaccum and solvent),
so called implicitly polarized charges (IPolQ) \cite{ipolq2013}. 
By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
we represent the IPolQ charges in the electronic continuum correction of the Lipid14 model
by increasing the scaling factor $f_q$ to $0.8$. 
\footnote{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  I use 0.80 as an approximate round value. The use of 0.78---.79 might be more appropriate, though.}
